Details of specific functions

### 2024-10-03 inst/gitbook/03_specific-tutorials.Rmd
Update 03_specific-tutorials.Rmd

>>>>

## Disparity and distances

There are two ways to use distances in `dispRity`, either with your input data being directly a distance matrix or with your disparity metric involving some kind of distance calculations.

### Disparity data is distance

If your disparity data is a distance matrix, you can use the option `is.dist = TRUE` in `dispRity` to make sure that all the operations done on your data take into account the fact that your disparity data has distance properties.
For example, if you bootstrap the data, this will automatically bootstrap both rows AND columns (i.e. so that the bootstrapped matrices are still distances).
This also improves speed on some calculations if you use [disparity metrics](#disparity-metrics) directly implemented in the package by avoiding recalculating distances (the full list can be seen in `?dispRity.metric` - they are usually the metrics with `dist` in their name).

### Disparity metric is distance

On the other hand if your data is not a distance matrix but you are using a metric that uses some kind of distance calculations, you can use the option `dist.helper` to greatly speed up calculations.
`dist.helper` can be either a pre-calculated distance matrix (or a list of distance matrices) or, better yet, a function to calculate distance matrices, like `stats::dist` or `vegan::vegdist`.
This option directly stores the distance matrix separately in the RAM and allows the disparity metric to directly access it at every disparity calculation iteration, making it much faster.
Note that if you provide a function for `dist.helper`, you can also provide any un-ambiguous optional argument to that function, for example `method = "euclidean"`.

If you use a disparity metric implemented in `dispRity`, the `dist.helper` option is correctly loaded onto the RAM regardless of the argument you provide (a matrix, a list of matrix or any function to calculate a distance matrix).
On the other hand, if you use your own function for the disparity metric, make sure that `dist.helper` exactly matches the internal distance calculation function.
For example if you use the already implemented `pairwise.dist` metric all the following options will be using `dist.helper` optimally:

```{r, eval = FALSE}
## Using the dist function from stats (specifying it comes from stats)
dispRity(my_data, metric = pairwise.dist, dist.helper = stats::dist)

## Using the dist function from vegdist function (without specifying its origin)
dispRity(my_data, metric = pairwise.dist, dist.helper = vegdist)

## Using some pre-calculated distance with a generic function
my_distance_matrix <- dist(my_distance_data)
dispRity(my_data, metric = pairwise.dist, dist.helper = my_distance_matrix)

## Using some pre-calculated distance with a user function defined elsewhere
my_distance_matrix <- my.personalised.function(my_distance_data)
dispRity(my_data, metric = pairwise.dist, dist.helper = my_distance_matrix)
```

However, if you use a homemade metric for calculating distances like this:

```{r, eval = FALSE}
## a personalised distance function
my.sum.of.dist <- function(matrix) {
  return(sum(dist(matrix)))
}
```

The `dist.helper` will only work if you specify the function using the same syntax as in the user function:

```{r, eval = FALSE}
## The following uses the helper correctly (as in saves a lot of calculation time)
dispRity(my_data, metric = my.sum.of.dist, dist.helper = dist)

## These ones however, work but don't use the dist.helper (don't save time)
## The dist.helper is not a function
dispRity(my_data, metric = my.sum.of.dist, dist.helper = dist(my_data))
## The dist.helper is not the correct function (should be dist)
dispRity(my_data, metric = my.sum.of.dist, dist.helper = vegdist)
## The dist.helper is not the correct function (should be just dist)
dispRity(my_data, metric = my.sum.of.dist, dist.helper = stats::dist)
```
